---
title: "Glittr stats"
format: 
  html
---

In this report you can find some general statistics about [Glittr.org](https://glittr.org). The plots and statistics created are amongst others used in the manuscript. Since [Glittr.org](https://glittr.org) is an ongoing project these statistics are updated weekly. 

## Set up the environment

This is required if you run this notebook locally. Loading required packages.

```{r}
#| output: false
library(httr2)
library(ggplot2)
library(dplyr)
library(ggbreak)
library(cowplot)
library(downlit)
library(xml2)
```

To run locally, create a file named `.env` and add your GitHub PAT (variable named `PAT` ) and google api key (named `GOOGLE_API_KEY`) in there, e.g.:

```{bash, eval=FALSE}
# this is an example, store it as .env:
export PAT="ghp_aRSRESCTZII20Lklser3H"
export GOOGLE_API_KEY="AjKSLE5SklxuRsxwPP8s0"
```

Now use your UNIX terminal to source this file to get the keys as objects:

```{bash, eval=FALSE}
source .env
```

In R, get environment variables as objects:

```{r}
pat <- Sys.getenv("PAT")
google_api_key <- Sys.getenv("GOOGLE_API_KEY")
```

Setting colors. These correspond to the category colours on glittr.org. 

```{r}
glittr_cols <- c(
  "Scripting and languages" =             "#3a86ff",
  "Computational methods and pipelines" = "#fb5607",
  "Omics analysis" =                      "#ff006e",
  "Reproducibility and data management" = "#ffbe0b",
  "Statistics and machine learning" =     "#8338ec",
  "Others" =                              "#000000")
```

## Parse repository data

Using the glittr.org REST API to get repository metadata, among which the stargazers, recency, category, license and tags. 

```{r}
#| code-fold: true

# get all repositories content as nested list
parsed <- request("https://glittr.org/api/repositories") |>
  req_perform() |>
  resp_body_json()

# extract relevant items as dataframe
repo_info_list <- lapply(parsed$data, function(x) data.frame(
  repo = x$name,
  author_name = x$author$name,
  stargazers = x$stargazers,
  recency = x$days_since_last_push,
  url = x$url,
  license = ifelse(is.null(x$license), "none", x$license),
  main_tag = x$tags[[1]]$name,
  main_category = x$tags[[1]]$category
))

repo_info <- do.call(rbind, repo_info_list)

# create a column with provider (either github or gitlab)
repo_info$provider <- ifelse(grepl("github", repo_info$url), "github", "gitlab")

# create a factor for categories for sorting
repo_info$main_category <- factor(repo_info$main_category,
                                  levels = names(glittr_cols))

# category table to keep order the same in the plots
cat_table <- table(category = repo_info$main_category)
cat_table <- sort(cat_table)
```

Number of repositories: `r nrow(repo_info)`

## Get contributors info

Using the GitHub REST API to get the number of contributors for each repository on glittr.org. This takes a few minutes, so if the contributors haven't changed, it will use a cached version. 

```{r}
#| code-fold: true

# take long time to run, so try to use cache results if no repos have been 
# added in the meantime

# check if data/n_contributors.rds exists
if(file.exists("data/n_contributors.rds")) {
  n_contributors <- readRDS("data/n_contributors.rds")
} else {
  n_contributors <- NULL
}

# get contributors info only from github repos
repo_info_gh <- repo_info[repo_info$provider == "github", ]

# get contributor info from github api if update is needed
if(!identical(sort(repo_info_gh$repo), sort(names(n_contributors)))) {
  dir.create("data", showWarnings = FALSE)
  n_contributors <- sapply(repo_info_gh$repo, function(x) {
    
    # get repo contributors
    resp <- request("https://api.github.com/repos/") |>
      req_url_path_append(x) |>
      req_url_path_append("contributors") |>
      req_url_query(per_page = 1) |>
      req_headers(
        Accept = "application/vnd.github+json",
        Authorization = paste("Bearer", pat),
        `X-GitHub-Api-Version` = "2022-11-28",
      ) |>
      req_perform() 
    
    link_url <- resp_link_url(resp, "last")
    if(is.null(link_url)) {
      return(1)
    } else {
      npages <- strsplit(link_url, "&page=")[[1]][2] |> as.numeric()
      return(npages)
    }
  })
  
  # overwrite rds file
  saveRDS(n_contributors, "data/n_contributors.rds")
}

repo_info_gh$contributors <- n_contributors[repo_info_gh$repo]

```

## Get country information

Here we get country information for all authors and organizations. It uses the free text specified at 'location'. Since this can be anything, we use the google REST API to translate that into country. 

```{r}
#| warning: false
#| output: false
#| code-fold: true

# check whether author info exists for caching
if(file.exists("data/author_info.rds")) {
  author_info <- readRDS("data/author_info.rds")
  author_info_authors <- unique(author_info$author) |> sort()
} else {
  author_info_authors <- NULL
}

gh_authors <- repo_info$author_name[repo_info$provider == "github"] |>
  unique() |>
  sort()

# if the author info is out of date, update it
if(!identical(gh_authors, author_info_authors)) {
  author_info_list <- list()
  for(author in gh_authors) {
    
    parsed <- request("https://api.github.com/users/") |>
      req_url_path_append(author) |>
      req_headers(
        Accept = "application/vnd.github+json",
        Authorization = paste("Bearer", pat),
        `X-GitHub-Api-Version` = "2022-11-28",
      ) |>
      req_perform() |>
      resp_body_json()
    
    author_info_list[[author]] <- data.frame(
      author = parsed$login,
      type = parsed$type,
      name = ifelse(is.null(parsed$name), NA, parsed$name),
      location = ifelse(is.null(parsed$location), NA, parsed$location)
    )
  }
  
  author_info <- do.call(rbind, author_info_list)
  
  author_info_loc <- author_info[!is.na(author_info$location), ]
  
  author_loc <- author_info_loc$location
  names(author_loc) <- author_info_loc$author
  
  ggmap::register_google(key = google_api_key)
  loc_info <- ggmap::geocode(author_loc,
                             output = 'all')
  
  get_country <- function(loc_results) {
    if("results" %in% names(loc_results)) {
      for(results in loc_results$results) {
        address_info <- results$address_components |> 
          lapply(unlist) |> 
          do.call(rbind, args = _) |>
          as.data.frame()
        country <- address_info$long_name[address_info$types1 == "country"]
        if (length(country) == 0) next
      }
      if (length(country) == 0) return(NA)
      return(country)
    } else {
      return(NA)
    }
  }
  
  countries <- sapply(loc_info, get_country)
  names(countries) <- names(author_loc)
  
  author_info$country <- countries[author_info$author]
  
  saveRDS(author_info, "data/author_info.rds")
}

repo_info <- merge(repo_info, author_info, by.x = "author_name",
                   by.y = "author")
repo_info$country[is.na(repo_info$country)] <- "undefined"
```

- Number of authors: `r nrow(author_info)`
- Number of countries: `r unique(repo_info$country) |> length()`

## Parse tag data

Here, we create `tag_df` that contains information for each tag by using the glittr.org API. 

```{r}
parsed <- request("https://glittr.org/api/tags") |>
  req_perform() |>
  resp_body_json()

tag_dfs <- list()
for(i in seq_along(parsed)) {
  category <- parsed[[i]]$category
  name <- sapply(parsed[[i]]$tags, function(x) x$name)
  repositories <- sapply(parsed[[i]]$tags, function(x) x$repositories)
  tag_dfs[[category]] <- data.frame(name, category, repositories)
}

tag_df <- do.call(rbind, tag_dfs) |> arrange(repositories)
```

Number of tags/topics: `r nrow(tag_df)`

## Number of repositories by category

This is figure 2A in the manuscript. 

```{r}
#| label: fig-categories
#| fig-cap: Number of repositories per category

cat_count_plot <- table(category = repo_info$main_category) |>
  as.data.frame() |>
  ggplot(aes(x = reorder(category, Freq), y = Freq, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = glittr_cols) +
  coord_flip() +
  theme_classic() +
  ggtitle("Categories") +
  theme(legend.position = "none",
        axis.title.y = element_blank()) +
  ylab("Number of repositories")

print(cat_count_plot)
```

And a table with the actual numbers 

```{r}
#| label: tbl-categories
#| tbl-cap: Number of repositories per category

category_count <- table(category = repo_info$main_category) |> as.data.frame()
knitr::kable(category_count)
```

## Number of contributors per repository separated by category

This is figure 2B in the manuscript. 

```{r}
#| label: fig-contributors
#| fig-cap: Number of contributors per repository separated by category

repo_info_gh$main_category <- factor(repo_info_gh$main_category,
                                     levels = names(cat_table))

contributors_plot <- repo_info_gh |>
  ggplot(aes(x = main_category, y = contributors, fill = main_category)) +
  geom_violin(scale = "width") +
  geom_boxplot(width = 0.1, col = "darkgrey") +
  coord_flip() +
  ggtitle("Contributors") +
  ylab("Number of contributors") +
  scale_y_sqrt() +
  scale_fill_manual(values = glittr_cols) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        plot.margin = margin(t = 5, r = 10, b = 5, l = 10))

print(contributors_plot)
```

And some statistics of contributors. 

```{r}
nna_contr <- repo_info_gh$contributors
param1 <- sum(nna_contr > 10)/length(nna_contr)
param2 <- sum(nna_contr > 1)/length(nna_contr)
param3 <- sum(nna_contr <= 5)/length(nna_contr)
```

- More than 10 contributors: `r signif(param1*100, digits = 3) `%
- More than 1 contributor: `r signif(param2*100, digits = 3) `%
- Between 1 and 5 contributors: `r signif(param3*100, digits = 3) `%

## Number of repositories per tag

This is figure 2C in the manuscript.

```{r}
#| label: fig-tags
#| fig-cap: Number of repostories per tag, colored by category.

tag_freq_plot <- tag_df |>
  filter(repositories > 10) |>
  ggplot(aes(x = reorder(name, repositories),
             y = repositories, fill = category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = glittr_cols) +
  ggtitle("Tags with > 10 repositories") +
  ylab("Number of repositories") +
  annotate(geom = "text", x = 2, y = 150,
           label = paste("Total number of tags: ",
                         nrow(tag_df)),
           color="black") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.y = element_blank())

print(tag_freq_plot)
```

And a table with the actual numbers.

```{r}
#| label: tbl-tags
#| tbl-cap: Number of repositories per tag

tag_df |>
  filter(repositories > 10) |>
  arrange(desc(repositories)) |>
  knitr::kable(row.names = FALSE)
```

## Number of repositories by author

This is figure 2D in the manuscript.

```{r}
#| label: fig-author
#| fig-cap: Number of repositories per author colored by category

author_freq <- table(author_name = repo_info$author_name, 
                     main_category = repo_info$main_category) |>
  as.data.frame()

author_freq$main_category <- factor(author_freq$main_category,
                                     levels = names(cat_table))

repos_per_author <- table(repo_info$author_name)

lf_authors <- names(repos_per_author)[repos_per_author < 5]

author_freq_plot <- author_freq |>
  filter(!author_name %in% lf_authors) |>
  arrange(Freq) |>
  ggplot(aes(x = reorder(author_name, Freq), y = Freq, fill = main_category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ggtitle("Author or organization") +
  ylab("Number of repositories") +
  scale_fill_manual(values = glittr_cols) +
  annotate(geom = "text", x = 2, y = 30,
           label = paste("Authors with < 5 repos: ",
                         length(lf_authors)),
           color="black") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.y = element_blank())

print(author_freq_plot)
```

And a table with the actual numbers.

```{r}
#| label: tbl-author
#| tbl-cap: Number of repositories per author

table(repo_info$author_name) |>
  as.data.frame() |>
  filter(Freq >= 5) |>
  arrange(desc(Freq)) |>
  knitr::kable()

```

## Number of repositories per license

This is figure 2E in the manuscript.

```{r}
#| label: fig-license
#| fig-cap: Number of repositories per license

lic_freq_data <- table(license = repo_info$license,
                       main_category = repo_info$main_category) |>
  as.data.frame()

lic_freq_data$main_category <- factor(lic_freq_data$main_category,
                                     levels = names(cat_table))

lic_freq_plot <- lic_freq_data |>
  ggplot(aes(x = reorder(license, Freq), y = Freq, fill = main_category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = glittr_cols) +
  theme_classic() +
  ggtitle("License type") +
  ylab("Number of repositories") +
  theme(legend.position = "none",
        axis.title.y = element_blank())

print(lic_freq_plot)
```

And a table with the actual numbers.

```{r}
#| label: tbl-license
#| tbl-cap: Number of repositories per license

repo_info$license |>
  table() |>
  as.data.frame() |>
  mutate(perc = round(Freq/nrow(repo_info)*100, 1)) |>
  arrange(desc(Freq)) |>
  knitr::kable()
```


## Number of repositories per country

This is figure 2F in the mansucript. 

```{r}
#| label: fig-country
#| fig-cap: Number of repositories per country colored by category
#| 
country_freq <- table(country = repo_info$country, 
                      main_category = repo_info$main_category) |>
  as.data.frame()

country_freq$main_category <- factor(country_freq$main_category,
                                     levels = names(cat_table))

country_freq_plot <- country_freq |>
  filter(country != "undefined") |>
  ggplot(aes(x = reorder(country, Freq), y = Freq, fill = main_category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ggtitle("Country") +
  ylab("Number of repositories") +
  scale_fill_manual(values = glittr_cols) +
  annotate(geom = "text", x = 2, y = 70,
           label = paste("Repos with undefined country: ",
                         sum(repo_info$country == "undefined")),
           color="black") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.y = element_blank())

print(country_freq_plot)
```

And a table with the actual numbers.

```{r}
#| label: tbl-country
#| tbl-cap: Number of repositories per country

repo_info$country |> 
  table() |> 
  as.data.frame() |> 
  arrange(desc(Freq)) |> 
  knitr::kable()
```


## Summary plot

Full figure 2 of the manuscript. 

```{r}
p <- plot_grid(cat_count_plot, contributors_plot, 
          tag_freq_plot, author_freq_plot,  
          lic_freq_plot, country_freq_plot,
          ncol = 2, labels = LETTERS[1:6],
          rel_heights = c(2,3,3))

ggsave("grid_plot_fig2.pdf", width = 10, height = 10)
ggsave("grid_plot_fig2.eps", width = 10, height = 10)
```






